---
title: "Coursera Data Science Capstone - Milestone Report"
author: "E. Manton"
date: "February 16, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

The **Coursera Data Science Capstone - Milestone Report** (aka, **"the report"**) is intended to give an introductory look at analyzing the SwiftKey data set and figuring out: 

1. What the data consists of, and 
2. Identifying the standard tools and models used for this type of data.

The report is then to be written in a clear, concise, style that a data scientist _OR_ non-data scientist can understand and make sense of.  

### Purpose

The purpose of the report is a four-fold **exploratory data analysis** that will:

1. Demonstrate that the data has been downloaded from Swiftkey (via Coursera) and successfully loaded into R.

2. Create a basic report of summary statistics about the data sets to include:
  + Word counts, line counts and basic data tables,
  + Basic plots, such as histograms and pie charts, to illustrate features of the data.

3. Report any interesting findings about the data so far amassed.

4. Present the basic plan behind creating a prediction algorithm and Shiny app from the data.

### Load Data, Calculate Size, Word Count, and Summarize

The Swiftkey data consists of four datasets, each in a different language (one each in German, English, Russian, and Finnish), containing random:

1. blog entries
2. news entries
3. twitter feeds

For this report, we will process the **English data**, and reference the German, Finnish, and Russian sets to possibly match foreign language characters and/or words embedded in the English data.

#### **Download the data** and unzip on Windows 10 PC
```{r unzip_data, eval=FALSE}
PkZip extract all... 
```

#### **Read the data into R** from the connection (simple text files) 
``` {r read_data}
blog_entries<-readLines("C:/Users/Mommy/OneDrive/Coursera/Data Science Specialization/Capstone/Data/final/en_US/en_US.blogs.txt", skipNul = TRUE, warn= FALSE)
news_entries<-readLines("C:/Users/Mommy/OneDrive/Coursera/Data Science Specialization/Capstone/Data/final/en_US/en_US.news.txt", skipNul = TRUE, warn=FALSE)
twitter_feeds<-readLines("C:/Users/Mommy/OneDrive/Coursera/Data Science Specialization/Capstone/Data/final/en_US/en_US.twitter.txt", skipNul = TRUE, warn=FALSE)
```

#### Calculate the **SIZE** of the English dataset (in megabytes) and display
``` {r file_size}
blog_entries_size<-file.info("C:/Users/Mommy/OneDrive/Coursera/Data Science Specialization/Capstone/Data/final/en_US/en_US.blogs.txt")$size/ 1024 ^ 2
news_entries_size<-file.info("C:/Users/Mommy/OneDrive/Coursera/Data Science Specialization/Capstone/Data/final/en_US/en_US.news.txt")$size/ 1024 ^ 2
twitter_feeds_size<-file.info("C:/Users/Mommy/OneDrive/Coursera/Data Science Specialization/Capstone/Data/final/en_US/en_US.twitter.txt")$size/ 1024 ^ 2
```

``` {r data_frame_size}
eng_data_set_size<-c(blog_entries_size,news_entries_size,twitter_feeds_size)
data_frame_size<-data.frame(eng_data_set_size)
names(data_frame_size)[1] <-"MBs"
row.names(data_frame_size) <- c("Blog entries", "News entries", "Twitter Feeds")
data_frame_size
```

#### Calculate the **LINE COUNT** of the English dataset and display
```{r data_frame_line_count}
blog_entries_line_count<-length(blog_entries)
news_entries_line_count<-length(news_entries)
twitter_feeds_line_count<-length(twitter_feeds)

data_set_length <-c(blog_entries_line_count,news_entries_line_count, twitter_feeds_line_count)
eng_data_frame_line_count <-data.frame(data_set_length)
names(eng_data_frame_line_count)[1] <-"Line Count"
row.names(eng_data_frame_line_count) <- c("Blog entries", "News entries", "Twitter Feeds")
eng_data_frame_line_count
```

#### Calculate the **WORD COUNT** of the English dataset and display
```{r word_count}
library(ngram)
blog_entries_word_count <-wordcount(blog_entries)
news_entries_word_count <-wordcount(news_entries)
twitter_feeds_word_count <-wordcount(twitter_feeds)

data_set_word_count <-c(blog_entries_word_count, news_entries_word_count, twitter_feeds_word_count)
eng_data_frame_word_count <-data.frame(data_set_word_count)
names(eng_data_frame_word_count)[1] <-"Word Count"
row.names(eng_data_frame_word_count) <- c("Blog entries", "News entries", "Twitter Feeds")
eng_data_frame_word_count
```

#### **SUMMERIZE** the English dataset and display two sample entries from each file
```{r summaries}
summary(blog_entries)
head(blog_entries,2)
summary(news_entries)
head(news_entries,2)
summary(twitter_feeds)
head(twitter_feeds,2)
```

### Basic Plots
Because of the amount of data that needs processing, and because this is an **exploratory data analysis**, we will extract the seventy-five most frequently used words from within each data file, and then proceed with some basic plotting.

#### Reduced Sample Size - training sets
We will use **1/100th** of each data file for our reduced sample size, and create the necessary data subsets from them.

```{r sample_size}
sample_size <- 0.01
blogs_index <- sample(seq_len(blog_entries_line_count),blog_entries_line_count*sample_size)
news_index <- sample(seq_len(length(news_entries)),length(news_entries)*sample_size)
twitter_index <- sample(seq_len(length(twitter_feeds)),length(twitter_feeds)*sample_size)

blogs_sub <- blog_entries[blogs_index[]]
news_sub <- news_entries[news_index[]]
twitter_sub <- twitter_feeds[twitter_index[]]
```
#### Combine the Subsets - final training set
We will now create a corpus from the data subsets.  The **tm** Library will assist us in this task.  The process involves removing all non-ASCII character data, punctuation marks, excess white space, numeric data, converting the remaining alpha characters to lower case, and generating the entire corpus in plain text.  A brief summary of the corpus is provided.

```{r korpus}
library(tm)
korpus <- Corpus(VectorSource(c(blogs_sub, news_sub, twitter_sub)),readerControl=list(reader=readPlain,language="en"))
korpus <- Corpus(VectorSource(sapply(korpus, function(row) iconv(row, "latin1", "ASCII", sub=""))))
korpus <- tm_map(korpus, removePunctuation)
korpus <- tm_map(korpus, stripWhitespace) 
korpus <- tm_map(korpus, content_transformer(tolower)) 
korpus <- tm_map(korpus, removeNumbers) 
korpus <- tm_map(korpus, PlainTextDocument) 
korpus <- Corpus(VectorSource(korpus))
head(korpus,5)
```

#### Display as a **HISTOGRAM**
We will now use our cleaned data subset to generate a histogram of the thirty most frequently used words in the corpus.  The libraries **slam** and **ggplot2** will help with this task.

```{r histogram}
library(slam)
library(ggplot2)
s_korpus <-TermDocumentMatrix(korpus,control=list(minWordLength=1))
wordFrequency <-rowapply_simple_triplet_matrix(s_korpus,sum)
wordFrequency <-wordFrequency[order(wordFrequency,decreasing=T)]
mostFrequent30 <-as.data.frame(wordFrequency[1:30])
mostFrequent30 <-data.frame(Words = row.names(mostFrequent30),mostFrequent30)
names(mostFrequent30)[2] = "Frequency"
row.names(mostFrequent30) <-NULL
mf30Plot = ggplot(data=mostFrequent30, aes(x=Words, y=Frequency, fill=Frequency)) + geom_bar(stat="identity") +  guides(fill=FALSE) + theme(axis.text.x=element_text(angle=90))
# mf30Plot +labs(title="30 Most Frequently Used Words")
mf30Plot + ggtitle("30 Most Frequently Used Words") + theme(plot.title = element_text(hjust = 0.5))
```

#### Display as a **PIE CHART**
We will now use our cleaned data subset to generate a pie chart of the five most frequently used words in the corpus.  The library **plotrix** will help with this task.

```{r scatter_plot}
library(plotrix)
mostFrequent10 <-head(mostFrequent30,5)
pie3D(mostFrequent10$Frequency, labels = mostFrequent10$Words, main = "Pie of Five Greatest Word Frequencies", explode=0.1, radius=1.8, labelcex = 1.3, start=0.7)
```

### Interesting Findings
With the exploratory data analysis done on the English data set to this point, the findings regarding the top 30 most frequently occurring words, to include the five most frequently occurring words, are not that surprising.  The bulk of them are articles and pronouns.  Further analysis using bigrams and trigrams would give better most frequently used phrase distribution.  This type of finding could then be used to predict trends in the data and to create a predictive model of English text.

### Basic Project Plan
The basic plan is to use the initial data analysis presented herein to further progress with the prediction algorithm necessary for the Shiny application - a predictive model of English text.  One way of doing this might be to investigate what is possible using Markov Chains.  Further analysis will be done using NGram modeling, to predict next-word selection with accuracy.  All will be incorporated into a user-friendly Shiny front end that will allow the user to interact with the data and make logical next-word selections.
